GEEPERs: Principal Stratification using Principal Scores and Stacked Estimating Equations

### 2025-07-01 writeUps/paper.tex

\begin{ass}[Residualized Covariate Independence]\label{ass:rci}
There exists $\bxy$, a (known) transformation of covariates $\bx$, and an (unknown) vector of coefficients $\bm{\gamma}$ such that
\begin{equation}\label{eq:ass:rci}
\begin{split}
\EE[\yc-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yc-\bm{\gamma}'\bxy|\st]\\
\EE[\yt-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yt-\bm{\gamma}'\bxy|\st]
\end{split}
\end{equation}
\end{ass}
That is, CI may not hold for potential outcomes themselves, but an analogous assumption holds for \emph{residualized} potential outcomes---i.e., $\yc$ and $\yt$ after subtracting out a linear function of (possibly transformed) covariates.
By forcing $\bm{\gamma}$ to be the same for both treatment groups, and (implicitly) for both principal strata, Assumption \ref{ass:rci} assumes that there are no interactions between columns of $\bxy$ and either condition or principal stratum.
This is akin to the ``Additivity of Treatment Assignment Effect'' assumption in \citet{jo2002}.
Proposition \ref{prop:interactions} in the appendix relaxes these requirements, but some preliminary simulations suggest that standard errors from estimates allowing for these interactions will be prohibitively large.
The simulation results in the following section address cases in which interactions are present in the data-generating model but not in the data analysis.

@@

We conducted a simulation study to compare the performance of \textsc{geepers} with that of a Bayesian mixture model and a principal score weighting estimator, in both favorable and unfavorable circumstances.
The study was designed to answer three sets of overarching questions.
First, how does the performance of the \textsc{geepers} estimator vary under different conditions, including sample size, the extent to which $\bx$ predicts $\st$, %whether and how the principal effects themselves vary, 
and when important interactions are omitted from the outcome model?
Second, how does \textsc{geepers} compare with mixture modeling? Is it competitive in circumstances favorable to both techniques? Does it avoid the pitfalls of mixture modeling when the parametric assumptions of the normal mixture model fail?
Third, are there ways or circumstances when the principal score weighting estimator outperforms \textsc{geepers} or a mixture estimator even though principal ignorability does not hold?

\subsection{Simulation Design}
The simulation study was conducted in \texttt{R} \citep{rcite} and Stan \citep{rstan}, and full replication code is available at %[Redacted]. %
\url{https://github.com/adamSales/psGee}.
We replicated each set of conditions 500 times.

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$, $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e., were included in the analysis model.
Given all three covariates, the principal scores were set as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps} was violated, i.e., $Var(\pp)=0$.
Principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$.

@@

\subsubsection{Manipulated Factors}\label{sec:manipulatedFactors}
We manipulated five factors in the simulation, listed in Table \ref{tab:factor}.
The five factors are not completely crossed---in particular, we let $n$ and $\alpha$ each vary across all the levels listed while holding the other four factors fixed.

\begin{table}
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|rl}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100,200,\dots,900,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$0,0.2,0.3,\dots,0.9,1$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, Uniform}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}} & None & $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& $\bm{x}:\st$ & $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& $\bm{x}:Z$&$\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& $\bm{x}:\st$ \& $\bm{x}:Z$& $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\end{tabular}
\end{table}

@@

$\bm{\alpha}$\textbf{:} As mentioned above, $\alpha$ controls the variance of the principal scores, i.e., the extent to which $\st$ is predictable as a function of covariates. When $\alpha=0$, the covariates are unrelated to $\st$ and principal scores do not vary between participants, violating \eqref{ass:vps}. We expect that the more predictive the covariates, the better the performance of all three estimators.
To aid in the interpretation of the $\alpha$ parameter, the online appendix plots $\alpha$ against the area under the receiver operating characteristic curve for each fitted model (AUC), a common measure of classification accuracy \citep{bradley1997use}---an AUC of 0.5 implies that the model is no better than random guessing, and AUC of 1 implies perfect prediction. For $0.5\le \alpha \le 1$, the average AUC varies from roughly 0.53 to 0.77; for $\alpha=0.5$, the average AUC was roughly 0.675.

\textbf{\emph{The distribution of $\bm{\epsilon_i}$:}} $\epsilon$ is the regression error in \eqref{eq:y-sim}. Across simulation runs, $E[\epsilon]=0$ and $Var(\epsilon)=1/2$. However, the shape of $\epsilon$'s distribution varied between the following two possibilities:
Normal $\epsilon\sim\mathcal{N}(0,1/2)$, as assumed by the mixture model estimator, or Uniform $\epsilon\sim\mathcal{U}(-\sqrt{6}/2,\sqrt{6}/2)$.
%\item[Mixture] $f_\epsilon(x)=\frac{1}{4}\phi\left((x+1/3)\sqrt{6}\right)+\frac{1}{4}\phi\left((x-1)\sqrt{6})\right)$
%The mixture distribution for $\epsilon$ was designed to be particularly difficult for the mixture model estimator.

@@

\textbf{\emph{Interactions between $\st$ or $Z$ and covariates:}} The factor---the presence of interactions between covariates and $\st$ and $Z$, respectively--- tests the performance of a model that does not include interactions when that modeling assumption is false.

@@

Using each simulated dataset, we estimated $\eff0$ and $\eff1$ with three methods: \geepers, a parametric mixture model \eqref{eq:mixtureModelPS} including principal scores---hereafter \pmm---and the principal score weighting estimator \eqref{eq:psw}, hereafter \psw.

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bxs=\bxy=\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data-generating model, $x_3$, was ``unobserved''--since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability \eqref{eq:pi} was violated.

@@

\pmm, the parametric mixture estimator, then fits two outcome models:
For the treatment group, the regression model $Y=\beta_0^T+\beta_1^TS+\bm{\gamma}'\bx+\epsilon$ and for the control group, a mixture model of the form \eqref{eq:mixtureModelPS}, but with linear normal models for $g^s(y|x_1,x_2)$. 
Residual variance was allowed to differ across treatment groups, but not across principal strata.
These models were fit simultaneously using Bayesian Markov Chain Monte Carlo with Stan \citep{rstan}. We interpreted the posterior mean as a point estimate and the posterior standard deviation as a standard error.

@@

\subsection{Results}\label{sec:simResults}
In this section, we evaluate estimates of $\eff1$. Results for estimates of $\eff0$ and $\eff1-\eff0$, which are similar, can be found in the online appendix. 
In replication $b=1,\dots,500$, define the error of a given estimator $\heff1_b$ as $err_b=\heff1_b-\eff1$. The bias of $\heff1$, $\EE[\heff1-\eff1]$, is estimated as $\widehat{bias}=\sum_b err_b/500$, and the empirical standard error is defined as $\left\{\sum_b(\heff1_b-\overline{\heff1})^2/499\right\}^{1/2}$, the sample standard deviation of $\heff1$. We also report the empirical coverage of confidence intervals, the proportion of replications in which $\eff1$ is within the estimated confidence bounds of $\heff1$. Here, we present results on error, bias, empirical standard error, and confidence interval coverage; results on root-mean-squared-error %$\left(\widehat{bias}^2+SE_{emp}^2\right)^{1/2}$ 
$\left(\sum_b err_b^2/500\right)^{1/2}$
are available in the online appendix.

\subsubsection{Varying $n$ and $\alpha$}
% When $\alpha=0$, Assumption \ref{ass:vps} was violated, and \textsc{geepers} estimates were highly erratic; however, as Table \ref{tab:coverage} shows, when other assumptions held, inference remained valid.
Our first set of results illustrates the performance of \geepers when all of its assumptions are met.
Since \geepers is an asymptotic method whose identification depends on $\bm{x}$ as a predictor of $\st$, we examine its bias and standard error as functions of $n$ and $\alpha$.
We include analogous results for \pmm as a baseline comparison.

Figure \ref{fig:alphan} shows the empirical bias and standard error (SE) of the \textsc{geepers} and \pmm estimators of $\eff1$ as sample size and $\alpha$ varied.
Across $n$ and $\alpha$, \textsc{geepers} had lower bias and higher SE than the mixture estimator.
\textsc{geepers} was positively biased for low $n$ with bias converging to close to 0 (up to simulation error) for $n\ge 300$, and was roughly unbiased (up to simulation error) for $\alpha \ge 0.2$, with $n$ fixed at 500.
The SE of the \textsc{geepers} decreased rapidly for $100\le n \le 500$ and $0.2\le \alpha \le 0.5$ and more gradually for higher values of $n$ and $\alpha$.

For the remainder of the simulation, we focus on the moderate cases of $n=500$ and $\alpha=0.2$ or 0.5.

@@

%<<<<<<< HEAD
Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation error for \geepers, the parametric mixture model \pmm, and the \psw estimator of $\eff1$. Table \ref{tab:coverage} shows empirical coverage of nominal 95\% confidence interval estimates for \geepers and \pmm; when at least one of an estimator's assumptions is violated, the result is colored red.

The first column of Figure \ref{fig:boxplots} and the black entries of Table \ref{tab:coverage} show results from scenarios in which all assumptions of \geepers and \pmm are met (\psw relies on principal ignorability, which is violated throughout the simulation study).
\textsc{geepers} is slightly less biased than the parametric mixture model; when $\alpha=0.2$, \textsc{geepers} is much more variable and $\alpha=0.5$ the difference in variance is slight.
\textsc{psw} has by far the lowest variance and the highest bias; the latter is due to the violation of principal ignorability. 
When all assumptions are met, confidence intervals from both \geepers and \pmm achieve their nominal levels.

The remainder of the results reflect violations modeling assumptions for either \geepers or the parametric mixture model or both: when $\alpha=0$, violating Assumption \ref{ass:vps}; interactions in the data generating model between covariates $X$ and treatment assignment $Z$ and/or principal stratum $S$, either of which would violate assumption \ref{ass:rci} and the linear additive model of the \textsc{pmm}; and/or residuals drawn from a uniform distribution, which violates the normality assumption of \textsc{pmm}.

We omitted the $\alpha=0$ case from Figures \ref{fig:alphan} and \ref{fig:boxplots} since the variance of \geepers and \pmm estimates was so large as to dominate the plot. 
However, the first column of results in Table \ref{tab:coverage} shows that, when other assumptions are met, confidence intervals from both methods tend to over-cover when $\alpha=0$. 
In a randomized experiment, $\alpha$ can be estimated directly, making Assumption \ref{ass:vps} testable.

For \geepers, when $\alpha=0.2$ the presence of interactions in the data-generating model increases sampling variance, but only interactions between $X$ and $Z$ reduce confidence interval coverage. When $\alpha=0.5$, the presence of interactions has little to no effect on the performance of the \geepers estimator, and all 95\% confidence intervals approximately achieve their nominal levels. \geepers makes no assumptions about the shape of the residual distribution and behaves similarly under either normal or uniform residuals.

Like \geepers, \pmm is insensitive to interactions between $X$ and $S$. However, interactions between $X$ and $Z$ in the data-generating model skew the \pmm sampling distribution and reduce the coverage of its 95\% confidence intervals, even when $\alpha=0.5$.
When the residual distribution was uniformly distributed, the \pmm sampling distribution appeared bimodal, and its credible intervals exhibited severe undercoverage.

\label{simsum}In sum, \psw is the least variable of all three estimators, and does not depend on outcome modeling, but is vulnerable to bias from an unmeasured covariate that predicts both $S$ and $Y$.
When all assumptions of both methods are met, \geepers is less efficient, but also less biased, than \pmm, but differences in efficiency decrease with $n$ and $\alpha$.
Neither \psw nor \pmm is vulnerable to interactions between $X$ and $S$.
\pmm is highly vulnerable to violations of normality and to interactions between $X$ and $Z$, while \psw is robust to non-normality and robust to $X$-$Z$ interactions when $\alpha = 0.5$.
These results suggest that researchers who are willing to assume principal ignorability should use \psw, and those who are confident in their parametric model for potential outcomes should use a parametric mixture model. Researchers who are hesitant to accept principal ignorability or a parametric outcome model should consider \geepers.

\label{evalMetric}The simulation study suggests that none of the three methods evaluated---\geepers, \pmm, or \psw---dominates the others, and offers some guidance on when each estimator may be preferable. \psw is the most precise and requires no assumptions about the distribution of outcomes, but can be quite sensitive to violations of principal ignorability. When the determinants of $\st$ are well-understood and measured, \psw seems to be the best option. When principal ignorability does not hold, potential outcomes are conditionally normal, and there are no $X:Z$ interactions within principal strata, the \pmm estimator outperforms \geepers, though the difference decreases with $n$ or $\alpha$. \geepers may be the best choice when conditional normality may not hold, or if there are $X:Z$ interactions and $\bm{x}$ is sufficiently predictive of $\st$ (AUC $\ge$0.675, corresponding to $\alpha \ge 0.5$, may serve as a threshold). 
These suggestions may depend heavily on the simulation setup; applying them in general scenarios requires further research.

@@

\subsection{Accounting for Imperfect Compliance: Re-Analysis of the Obstetrics and Periodontal Therapy Study}
\sloppy
Periodontal disease, a weakening of bone and connective tissue supporting teeth, is highly prevalent among American adults and is associated with adverse pregnancy outcomes, including preterm birth \citep{periodontalEpi}.
The Obstetrics and Periodontal Therapy Study \citep[(OPTS)][]{michalowicz2006treatment} investigated the efficacy of nonsurgical periodontal treatment to reduce the preterm birth rate; replication data are available at \url{https://causeweb.org/tshs/obstetrics-and-periodontal-therapy/}.
Pregnant patients at four different hospitals who were at least 16 years old, at less than 16 weeks and 6 days of gestation, and diagnosed with periodontal disease were individually randomized to either receive periodontal treatment during the first 21 weeks of pregnancy ($Z=1$, $n_T=413$) or after delivery ($Z=0$, $n_C=410$).
The primary prespecified outcome of the study was the babies' gestational age at birth; however, the study team also collected data on several secondary birth outcomes, as well as measures of patients' periodontal disease at follow-up.

The main component of the intervention consisted of up to four visits, followed by monthly tooth polishing and other dental care as needed. However, not all participants randomized to the treatment condition completed the entire treatment plan.
However, 96\% of patients in the treatment group received at least some periodontal care during pregnancy---therefore, there is reason to suspect nonzero treatment effects even for those patients who did not (or would not) fully comply.
Patients randomized to the control condition did not have access to the intervention.
\citet{richardson2023estimating} used a method they refer to as ``bespoke instrumental variable'' (BSIV) to estimate the average effects for the principal stratum of subjects who, if randomized to treatment,  would complete the treatment plan and for the principal stratum of subjects who would not.
Here, we will use \textsc{geepers}, a Bayesian mixture model, and principal score weighting to estimate the same quantities, following the same analysis decisions as \citet{richardson2023estimating}.
Replication code for our analysis can be found at \url{ https://github.com/adamSales/psGee}.

>>>>

\begin{assp}{\ref*{ass:ci}$'$}[Residualized Covariate Independence]\label{ass:rci}
There exists $\bxy$, a (known) multivariate transformation of covariates $\bx$, and an (unknown) vector of coefficients $\bm{\gamma}$ such that
\begin{equation}\label{eq:ass:rci}
\begin{split}
\EE[\yc-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yc-\bm{\gamma}'\bxy|\st]\\
\EE[\yt-\bm{\gamma}'\bxy|\bxs,\st]&=\EE[\yt-\bm{\gamma}'\bxy|\st]
\end{split}
\end{equation}
\end{assp}
That is, CI may not hold for potential outcomes themselves, but an analogous assumption holds for \emph{residualized} potential outcomes---i.e., $\yc$ and $\yt$ after subtracting out a linear function of (possibly transformed) covariates.
By forcing $\bm{\gamma}$ to be the same for both treatment groups, and (implicitly) for both principal strata, Assumption \ref{ass:rci} assumes that there are no interactions between columns of $\bxy$ and either condition or principal stratum.
This is akin to the ``Additivity of Treatment Assignment Effect'' assumption in \citet{jo2002}.
Proposition \ref{prop:interactions} in the appendix relaxes these requirements, but some preliminary simulations suggest that standard errors from estimates allowing for these interactions will be prohibitively large.
The simulation results in the following section address cases in which interactions are present in the data-generating model but not in the data analysis.

@@

We conducted a simulation study to compare the performance of \textsc{geepers} with that of a parametric mixture model (\pmm) and a principal score weighting (\psw) estimator, in both favorable and unfavorable circumstances.
The study was designed to answer three sets of overarching questions.
First, how does the performance of the \textsc{geepers} estimator vary under different conditions, including sample size, the extent to which $\bx$ predicts $\st$, %whether and how the principal effects themselves vary, 
and when important interactions are omitted from the outcome model?
Second, how does \textsc{geepers} compare with parametric mixture modeling? Is it competitive in circumstances favorable to both techniques? Does it avoid the pitfalls of \pmm when the parametric assumptions of the normal mixture model fail?
Third, are there ways or circumstances when \psw outperforms \textsc{geepers} or \pmm even though principal ignorability does not hold?

\subsection{Simulation Design}
The simulation study was conducted in \texttt{R} \citep{rcite} and Stan \citep{rstan}, and full replication code is available at [Redacted]. %
%\url{}.
We replicated each set of conditions 500 times.

@@

In each run of the simulation, we simulated three independent standard normal covariates, $x_k$, $k=1,\dots,3$. However, only the first two covariates were ``observed,'' i.e., were included in the analysis model.
Given all three covariates, the principal scores were set as
\begin{equation*}
  \ppi=logit^{-1}\left[\alpha(x_{1i}-x_{2i}+x_{3i})\right]
\end{equation*}
where $\alpha$ is a manipulated factor.
Principal stratum $\sti\in \{0,1\}$ was simulated as $S_i\sim Bern(\ppi)$.

@@

\subsubsection{Manipulated Factors}\label{sec:manipulatedFactors}
We manipulated four factors in the simulation, listed in Table \ref{tab:factor}.
The factors were completely crossed, except for some levels of $n$ and $\alpha$, which were tested with other factors held fixed.

% \begin{table}
%     \caption{\label{tab:factor} Manipulated factors in simulation study}
%   \centering
%   %\rowcolors{1}{}{lightgray}
% \begin{tabular}{r|ll}%{*{3}{c}}
%   \hline
%   Factor &\multicolumn{2}{l}{Levels}\\
%   & Part 1 & Part 2\\
%   \hline
% Sample size per condition $n$ &$100,200,\dots,1000$ & 500, 1000\\
% $\alpha$ & $0,0.2,0.3,\dots,1$ & 0, 0.2, 0.5 \\
% The distribution of $\epsilon_i$& Normal & Normal, Uniform\\
% %$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
% \multirow{8}{*}{Interactions in \eqref{eq:y-sim}} &  \multirow{8}{1.5in}{None $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$}& None\\
% & &$\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
% & & $\bm{x}:\st$\\
% && $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
% & & $\bm{x}:Z$\\
% &&$\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
% && $\bm{x}:\st$ \& $\bm{x}:Z$\\
% & & $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% % between $\st$ and covariates & No, Yes\\
% % Interactions between $Z$ and covariates& No, Yes\\
% \hline
% \end{tabular}
% \end{table}

\begin{table}
    \caption{\label{tab:factor} Manipulated factors in simulation study}
  \centering
  %\rowcolors{1}{}{lightgray}
\begin{tabular}{r|ll}%{*{3}{c}}
  \hline
  Factor &\multicolumn{2}{l}{Levels}\\
  \hline
Sample size per condition $n$ &\multicolumn{2}{l}{$100^*,200^*,300^*,400^*,500,600^*,700^*,800^*,900^*,1000$}\\
$\alpha$ &\multicolumn{2}{l}{$0,0.2,0.3^\dagger,0.4^\dagger,0.5,0.6^\dagger,0.7^\dagger,0.8^\dagger,0.9^\dagger,1^\dagger$}\\
The distribution of $\epsilon_i$&\multicolumn{2}{l}{ Normal, Uniform}\\
%$\beta_1$ & \multicolumn{2}{l}{0, 0.3}\\
\multirow{4}{*}{Interactions in \eqref{eq:y-sim}}& None:  $\gamma_1=1/\sqrt{6};\gamma_2=\gamma_3=0$\\ 
& $\bm{x}:\st$: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=1/(2\sqrt{6})$; $\gamma_3=0$\\
& $\bm{x}:Z$: $\gamma_1=1/\sqrt{6};\gamma_2=0$; $\gamma_3=1/(2\sqrt{6})$\\
& $\bm{x}:\st$ \& $\bm{x}:Z$: $\gamma_1=3/(4\sqrt{6})$; $\gamma_2=\gamma_3=1/(2\sqrt{6})$\\
% between $\st$ and covariates & No, Yes\\
% Interactions between $Z$ and covariates& No, Yes\\
\hline
\multicolumn{3}{p{4.5in}}{\footnotesize Factors fully crossed except where indicated.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $*$ Only tested with $\alpha=0.5$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}\\
\multicolumn{3}{p{4.5in}}{\footnotesize $\dagger$ Only tested with $n=500$, normal $\epsilon$, no interactions in \eqref{eq:y-sim}.}

\end{tabular}
\end{table}

@@

$\bm{\alpha}$\textbf{:} $\alpha$ controlled the extent to which $\st$ is predictable as a function of covariates. When $\alpha$ was higher, $\st$ was more easily predicted by covariates; when $\alpha=0$, Assumption \ref{ass:vps} was violated, i.e., $Var(\pp)=0$. We expected that the more predictive the covariates, the better the performance of all three estimators.
To aid in the interpretation of the $\alpha$ parameter, the online appendix plots $\alpha$ against the area under the receiver operating characteristic curve for each fitted model (AUC), a common measure of classification accuracy \citep{bradley1997use}---an AUC of 0.5 implies that the model is no better than random guessing, and AUC of 1 implies perfect prediction. For $0.5\le \alpha \le 1$, the average AUC varies from roughly 0.53 to 0.77; for $\alpha=0.5$, the average AUC was roughly 0.675.

\textbf{\emph{The distribution of $\bm{\epsilon_i}$:}} $\epsilon$ is the regression error in \eqref{eq:y-sim}. Across simulation runs, $E[\epsilon]=0$ and $Var(\epsilon)=1/2$. However, the shape of $\epsilon$'s distribution varied between the following two possibilities:
Normal $\epsilon\sim\mathcal{N}(0,1/2)$, as assumed by \pmm, or Uniform $\epsilon\sim\mathcal{U}(-\sqrt{6}/2,\sqrt{6}/2)$.
%\item[Mixture] $f_\epsilon(x)=\frac{1}{4}\phi\left((x+1/3)\sqrt{6}\right)+\frac{1}{4}\phi\left((x-1)\sqrt{6})\right)$
%The mixture distribution for $\epsilon$ was designed to be particularly difficult for the mixture model estimator.

@@

\textbf{\emph{Interactions between $\st$ or $Z$ and covariates:}} This factor was designed to test the vulnerability of the estimators to certain violations or Assumption \ref{ass:rci}.

@@

Using each simulated dataset, we estimated $\eff0$ and $\eff1$ with three methods: \geepers, \pmm with principal scores \eqref{eq:mixtureModelPS}, and the \psw estimator \eqref{eq:psw}.

The methods all used the same ``observed'' dataset, consisting of outcomes $Y$, treatment assignments $Z$, two covariates $\bx=[x_1,x_2]$, and $S$ observed only when $Z=1$.
The third covariate used in the data-generating model, $x_3$, was ``unobserved''---since $x_3$ was correlated with both $\st$ and $Y$, Principal Ignorability \eqref{eq:pi} was violated.

@@

\pmm, the parametric mixture estimator, then fits two outcome models:
For the treatment group, the regression model $Y=\beta_0^T+\beta_1^TS+\bm{\gamma}'\bx+\epsilon$ and for the control group, a mixture model of the form \eqref{eq:mixtureModelPS} with linear normal models for $g^s(y|x_1,x_2)$. 
Residual variance was allowed to differ across treatment groups, but not across principal strata.
These models were fit simultaneously using Bayesian Markov Chain Monte Carlo with Stan \citep{rstan}. We interpreted the posterior mean as a point estimate and the posterior standard deviation as a standard error.

@@

\subsection{Results}\label{sec:simResults}
In this section, we evaluate estimates of $\eff1$. Results for estimates of $\eff0$ and $\eff1-\eff0$, which are similar, can be found in the online appendix. 
In replication $b=1,\dots,500$, define the error of a given estimator $\heff1_b$ as $err_b=\heff1_b-\eff1$. The bias of $\heff1$, $\EE[\heff1-\eff1]$, is estimated as $\widehat{bias}=\sum_b err_b/500$, and the empirical standard error is defined as $\left\{\sum_b(\heff1_b-\overline{\heff1})^2/499\right\}^{1/2}$, the sample standard deviation of $\heff1$. We also report the empirical coverage of confidence intervals, the proportion of replications in which $\eff1$ is within the estimated confidence bounds of $\heff1$. Results on root-mean-squared-error %$\left(\widehat{bias}^2+SE_{emp}^2\right)^{1/2}$ 
$\left(\sum_b err_b^2/500\right)^{1/2}$
are in the online appendix.

\subsubsection{Performance with Varying $n$ and $\alpha$}
% When $\alpha=0$, Assumption \ref{ass:vps} was violated, and \textsc{geepers} estimates were highly erratic; however, as Table \ref{tab:coverage} shows, when other assumptions held, inference remained valid.
To what extent do the performances of \geepers and \psw depend on $n$ and $\alpha$?
For this set of results, $\epsilon\sim \mathcal{N}(0,1/2)$ and there were no interactions in \eqref{eq:y-sim}; $\alpha$ was fixed at $0.5$ for varying levels of $n$, and $n$ was fixed at 500 for varying levels of $\alpha$.

Figure \ref{fig:alphan} shows the empirical bias and standard error (SE) of the \textsc{geepers} and \pmm estimators of $\eff1$ as sample size and $\alpha$ varied.
Across $n$ and $\alpha$, \textsc{geepers} was approximately unbiased,\footnote{All p-values testing for \geepers bias were $>0.1$, with the exception of when $\alpha=1$, with $\widehat{bias}=0.01$, $p=0.046$ before adjustment for multiple comparisons.} while $\pmm$ estimators exhibited slight positive bias that decreased with $n$.

\geepers had a higher SE than \pmm, particularly when $n=100$ or $\alpha=0.2$. The \psw advantage is far less dramatic for higher levels of $n$ and $\alpha$, eventually diminishing to approximately zero.

@@

%<<<<<<< HEAD
Figure \ref{fig:boxplots} shows violin plots, layered on top of jittered scatter plots, showing estimation errors for \geepers, \pmm, and \psw. Table \ref{tab:coverage} shows empirical coverage rates of nominal 95\% confidence interval estimates for \geepers and \pmm; when at least one of an estimator's assumptions is violated, the result is colored red.

When all of the \geepers and \pmm assumptions were met, that is, when $\epsilon\sim\mathcal{N}(0,1/2)$, $\alpha>0$, and there were no covariate interactions in \eqref{eq:y-sim}, results for $\geepers$ and $\pmm$ were the same as in the previous section; in addition, confidence intervals from both methods achieved their nominal levels. 
\textsc{psw} had by far the lowest variance and the highest bias of the three estimators; the latter is due to the violation of principal ignorability.

%The remainder of the results reflect violations modeling assumptions for either \geepers or the parametric mixture model or both: when $\alpha=0$, violating Assumption \ref{ass:vps}; interactions in the data generating model between covariates $X$ and treatment assignment $Z$ and/or principal stratum $S$, either of which would violate assumption \ref{ass:rci} and the linear additive model of the \textsc{pmm}; and/or residuals drawn from a uniform distribution, which violates the normality assumption of \textsc{pmm}.

%We omitted the $\alpha=0$ case from Figures \ref{fig:alphan} and \ref{fig:boxplots} since the variance of \geepers and \pmm estimates was so large as to dominate the plot. 
When $\alpha=0$, Assumption \ref{ass:vps} is violated; %, standard errors for \geepers were too high to 
Nevertheless, as the first column of results in Table \ref{tab:coverage} shows, when other assumptions were met, confidence intervals from both \geepers and \psw tended to over-cover when $\alpha=0$. 
In a randomized experiment, $\alpha$ can be estimated directly, making Assumption \ref{ass:vps} testable.

The impact of covariate interactions in \eqref{eq:y-sim} on the performance of \geepers depended on $\alpha$: when $\alpha=0.2$, the presence of interactions increased sampling variance. Still, only interactions between $X$ and $Z$ reduced confidence interval coverage. When $\alpha=0.5$, interactions had little to no effect on the \geepers estimator, and all 95\% confidence intervals approximately achieved their nominal levels. \geepers performed similarly under either normal or uniform residuals.

Like \geepers, \pmm is insensitive to interactions between $X$ and $S$. However, interactions between $X$ and $Z$ in the data-generating model skewed its sampling distribution and reduced the coverage of its 95\% confidence intervals for all values of $\alpha$ in the simulation. 
Uniformly-distributed residuals caused bimodality in the \pmm sampling distribution, and caused its credible intervals to exhibit severe undercoverage.

\label{simsum}In sum, \psw is the least variable of all three estimators, and does not depend on outcome modeling, but is vulnerable to bias from an unmeasured covariate that predicts both $S$ and $Y$.
When all assumptions of both methods are met, \geepers is less efficient, but also less biased, than \pmm, but differences in efficiency decrease with $n$ and $\alpha$.
Neither \psw nor \pmm is vulnerable to interactions between $X$ and $S$.
\pmm is highly vulnerable to violations of normality and to interactions between $X$ and $Z$, while \psw is robust to non-normality and robust to $X$-$Z$ interactions when $\alpha = 0.5$.
These results suggest that researchers who are willing to assume principal ignorability should use \psw, and those who are confident in their parametric model for potential outcomes should use a parametric mixture model. Researchers who are hesitant to accept principal ignorability or a parametric outcome model should consider \geepers, particularly when $n\ge 200$ and covariates can predict $S$ with AUC$>0.6$ (corresponding to $\alpha>0.2$).
These suggestions may depend heavily on the simulation setup; applying them in general scenarios requires further research.

%\label{evalMetric}The simulation study suggests that none of the three methods evaluated---\geepers, \pmm, or \psw---dominates the others, and offers some guidance on when each estimator may be preferable. \psw is the most precise and requires no assumptions about the distribution of outcomes, but can be quite sensitive to violations of principal ignorability. When the determinants of $\st$ are well-understood and measured, \psw seems to be the best option. When principal ignorability does not hold, potential outcomes are conditionally normal, and there are no $X:Z$ interactions within principal strata, the \pmm estimator outperforms \geepers, though the difference decreases with $n$ or $\alpha$. \geepers may be the best choice when conditional normality may not hold, or if there are $X:Z$ interactions and $\bm{x}$ is sufficiently predictive of $\st$ (AUC $\ge$0.675, corresponding to $\alpha \ge 0.5$, may serve as a threshold).

@@

\subsection{Accounting for Imperfect Compliance: Re-Analysis of the Obstetrics and Periodontal Therapy Study}
\sloppy
Periodontal disease, a weakening of bone and connective tissue supporting teeth, is highly prevalent among American adults and is associated with adverse pregnancy outcomes, including preterm birth \citep{periodontalEpi}.
The Obstetrics and Periodontal Therapy Study \citep[OPTS; ][]{michalowicz2006treatment} investigated the efficacy of nonsurgical periodontal treatment to reduce the preterm birth rate; replication data are available at \url{https://causeweb.org/tshs/obstetrics-and-periodontal-therapy/}.
Pregnant patients at four different hospitals who were at least 16 years old, at less than 16 weeks and 6 days of gestation, and diagnosed with periodontal disease were individually randomized to either receive periodontal treatment during the first 21 weeks of pregnancy ($Z=1$, $n_T=413$) or after delivery ($Z=0$, $n_C=410$).
The main component of the intervention consisted of up to four visits, followed by monthly tooth polishing and other dental care as needed. Not all participants randomized to the treatment condition completed the entire treatment plan; however, 96\% of them received at least some periodontal care during pregnancy.
Patients randomized to the control condition did not have access to the intervention.

Following \citet{richardson2023estimating}, let $\st=1$ for patients who, if randomized to treatment, would complete the full treatment plan, and let $\st=0$ for those who would not.
Let $Y$, the outcome of interest, be the percentage of probed sites (6 sites per tooth) that were bleeding at visit 5, which was at 29--32 weeks of gestation, with potential values $\yc$ and $\yt$, as above.
\citet{richardson2023estimating} used a principal stratification estimator they called ``bespoke instrumental variables'' (BSIV) to estimate principal effects $\eff0$ and $\eff1$, the average effects of randomization to the treatment condition for patients who would or would not complete the full treatment plan.
In this section, we compare their result to estimates using \geepers, \pmm, and \psw.

For simplicity, we followed the same analysis decisions as \citet{richardson2023estimating}. 
Specifically, we conditioned on two covariates, Serum measures of endotoxin level and fibrinogen at baseline, and
omitted the 164 cases in which $Y$ was missing as well as 19 additional cases in which one or both of the covariates were missing.
That left a total sample size of 640, with 314 randomized to the treatment condition and 326 randomized to control.
%The compliance variable in the dataset had three categories: ``Yes,'' for subjects in the treatment group who completed the treatment plan (157 subjects), ``No'' for those who did not (4 subjects), and ``Und'' for those who received some therapy, but it is unknown whether they completed the treatment (157 subjects).
%We let $S=1$ when compliance was ``yes''---i.e., when subjects in the treatment group completed the treatment---and $S=0$ otherwise.
